---
title: Extreme weather events that caused the greatest harm to the population and
  the economy of the U.S. in 1996-2011
author: "Anton"
date: "Sunday, August 24, 2014"
output: pdf_document
---
#Synopsis  
The basic goal of this report is to explore the NOAA Storm Database and answer 2 basic questions about severe weather events across the United States:  
1. Which types of events are most harmful with respect to population health?  
2. Which types of events have the greatest economic consequences?  
To find out answers, data from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database is used. The data recieved from the National Weather Service that collected it from a variety of sources: county, state and federal emergency management officials, local law enforcement officials, skywarn spotters, NWS
damage surveys, newspaper clipping services, the insurance industry and the general public.  
The results show us that most harmful weather events are tornados (in case 1) and floods (in case2).      

#Data Processing    
The data downloaded from the [Storm Data sourse](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2). The events in the data start in the year 1950 and end in November 2011.  
```{r, cache = TRUE}
rawdata <- read.csv(bzfile("repdata-data-StormData.csv.bz2"), header = TRUE, 
                    stringsAsFactors = FALSE, strip.white=TRUE)
```
It's huge amount of the data. So now we need to reduce the data for our analyses. 
```{r}
names(rawdata)
```
According to the NOAA documentation all 48 weather event types recorded from 1996 to present. Now we can choose time range (no earlier than 1996/01/01) and required fields (date, event type, number of fatalities and injuries, property and crop damage estimates with their magnitude).  
```{r}
tidydata <- subset(rawdata, as.Date(rawdata$BGN_DATE, "%m/%d/%Y") >= "1996-01-01")
names(tidydata) <- tolower(gsub("_", "", names(tidydata)))
tidydata <- tidydata[c("bgndate", "evtype", "fatalities", "injuries",
                       "propdmg", "propdmgexp", "cropdmg", "cropdmgexp")]
tidydata$evtype <- tolower(tidydata$evtype)
```
Unfortunately event type parameter does not match the description in documentation. According to the documentaton we should have 48 unique values. But we have:
```{r}
length(unique(tidydata$evtype))
```
There are not enough information in the data to decide which types is "flood/flash/flood" or "winter mix", for example.
But we can choose some simple cases which are fully consistent with the documentation, for example: 
```{r}
tidydata$evtype[which(grepl("tstm wind$", tidydata$evtype))] <- "thunderstorm wind"
tidydata$evtype[which(grepl("tornado", tidydata$evtype))] <- "tornado"
tidydata$evtype[which(grepl("surge", tidydata$evtype))] <- "storm surge/tide"
```  
Now we can create two datasets in accordance with our objectives.   
```{r}
healthdmg <- tidydata[!tidydata$fatalities == 0 | !tidydata$injuries == 0,
                      c("evtype", "fatalities", "injuries")]

economicdmg <- tidydata[!tidydata$propdmg == 0 | !tidydata$cropdmg == 0,
                      c("evtype", "propdmg", "propdmgexp", "cropdmg", "cropdmgexp")]
```  
Also to answer the second question we need to convert numeric and character values into a numeric data.  

```{r, cache = TRUE}
strvalue <- c("K", "M", "B", "")
numvalue <- c(1000, 1000000, 1000000000, 1)
for (i in 1:nrow(economicdmg)) {
        a <- numvalue[strvalue == economicdmg$propdmgexp[i]]*economicdmg$propdmg[i]
        b <- numvalue[strvalue == economicdmg$cropdmgexp[i]]*economicdmg$cropdmg[i]
        economicdmg$dmg[i] <- a+b
}
```  
Finally we summarise and sort the data.   
```{r}
totalhealthdmg <- aggregate(fatalities+injuries~evtype, data=healthdmg, sum)
names(totalhealthdmg) <- c("evtype", "dmg")
totalhealthdmg <- totalhealthdmg[order(-totalhealthdmg$dmg),]
totalhealthdmg$evtypes <- reorder(totalhealthdmg$evtype, -totalhealthdmg$dmg)

totaleconomicdmg <- aggregate(dmg~evtype, data=economicdmg, sum)
totaleconomicdmg <- totaleconomicdmg[order(-totaleconomicdmg$dmg),]
totaleconomicdmg$evtypes <- reorder(totaleconomicdmg$evtype, -totaleconomicdmg$dmg)
```  

#Results    
Now we make two plots to show answers on our two questions.  
```{r}
library(ggplot2)
library(RColorBrewer)

cols <- brewer.pal(9, "Reds")
pal <- colorRampPalette(cols)
qplot(evtypes, dmg, data = totalhealthdmg[1:10,], 
      xlab = "Weather event type (top 10)", 
      ylab = "Total number of fatalities and injuries",
      geom = "histogram", fill = evtypes, stat = "identity")+
        scale_fill_manual(values=rev(pal(10)))+ theme_bw()+
        theme(axis.text.x = element_blank())


qplot(evtypes, dmg/1000000000, data = totaleconomicdmg[1:10,], 
      xlab = "Weather event type (top 10)", 
      ylab = "Total amount of damages (billions $)",
      geom = "histogram", fill = evtypes, stat = "identity")+
        scale_fill_brewer(palette="RdYlGn")+ theme_bw()+
        theme(axis.text.x = element_blank())
```    

In conclusion, we can see that tornados are most harmful with respect to population health while floods are the most destructive.

